Embracing the Laws of Physics: \\ Three Reversible Models of Computation

### 2017-12-28 survey/survey.tex
data I small changes

Before we start our investigation of what it means to have typed,
information preserving transformations, let's begin with just
finite sets. Most programming languages provide some collection of primitive
data like booleans, characters, strings, and (bounded) numbers that are
naturally modeled as sets. Infinite sets are much more subtle, and will be
discussed in a later section.

What does it mean for two finite sets to be ``the same''?  Well, clearly the
sets $A = \left\{1, 2, 3\right\}$ and $B = \left\{c, d\right\}$ are different.
Why?  Well, supposed there was a transformation $f : A \rightarrow B$ that
deformed $A$ into $B$, and another $g : B \rightarrow A$ which undid this
transformation. Since $f$ is total, by the pigeonhole principle, two elements
of $A$ would be mapped to the same element of $B$. Supposed that this is $2$ and
$3$, and that they both map to $d$.  But $g(d)$ cannot be both $2$ and $3$, and
so $g$ is not the inverse of $f$. With just a little more work, we can show
that $f$ (and $g$) must be both injective and surjective. In other words,
$f$ (and $g$) must be a bijection between $A$ and $B$.  And of course this 
only happens when $A$ and $B$ have the same number of elements.

More importantly, given a bijection $f : C \rightarrow D$ of finite sets
$C,D$, there always exists another bijection $g : D \rightarrow C$ which is
$f$'s inverse. So, for finite sets, \emph{bijections} act as reversible
deformations.

@@

Bijections on \fin{n} have a specific name: permutations. As is well-known
permutations can be generated by sequences of transpositions. Thus we
can create a small language for writing permutations as
\begin{verbatim}
p ::= id | swap i j | p;p
\end{verbatim}
where $i,j:\Nat$, $i\neq j$ and $i,j < n$. Note that we could remove 
\verb|id|
from the language and drop the $i\neq j$ condition so that 
\verb|swap j j| would represent the identity permutation.

While we are mainly interested in typed data, it is always worthwhile to
first start investigating the untyped setting first, as it gives us insights
into what the \emph{purely operational} view of the theory would be.
This tells us that permutations are an inescapable part of the fabric of
reversible computing.

However as permutations are untyped, and act on the canonicalized version
of $n$-element sets (i.e. those sets where all the structure has been
forgotten), these are a rather pale shadow of the rich tapestry of
information-perserving transformation of structured data.

>>>>

Some of the simplest spaces are the finite and discrete spaces of
points. Indeed most programming languages provide some collection of
primitive data like booleans, characters, strings, and (bounded)
numbers that are naturally modeled as finite sets. Infinite sets are much
more subtle, and will be discussed in the conclusion.

What does it mean for two finite sets to be ``the same''?  Well,
clearly the sets $A = \left\{1, 2, 3\right\}$ and
$B = \left\{c, d\right\}$ are different.  Why?  Well, supposed there
was a transformation $f : A \rightarrow B$ that deformed $A$ into $B$,
and another $g : B \rightarrow A$ which undid this
transformation. Since $f$ is total, by the pigeonhole principle, two
elements of $A$ would be mapped to the same element of $B$. Supposed
that this is $2$ and $3$, and that they both map to $d$.  But $g(d)$
cannot be both $2$ and $3$, and so $g$ is not the inverse of $f$. With
just a little more work, we can show that $f$ (and $g$) must be both
injective and surjective. In other words, $f$ (and $g$) must be a
bijection between $A$ and $B$.  And of course this only happens when
$A$ and $B$ have the same number of elements. More importantly, given
a bijection $f : C \rightarrow D$ of finite sets $C,D$, there always
exists another bijection $g : D \rightarrow C$ which is $f$'s
inverse. So, for finite sets, \emph{bijections} act as reversible
deformations.

@@

Bijections on \fin{n} have a specific name: permutations. As is well-known
permutations can be generated by sequences of transpositions. Thus we
can create a small language for writing permutations as
\[\begin{array}{rcl}
p &::=& \mathit{id} ~\mid~ \mathit{swap}~i~j ~\mid~ p;p
  \end{array}\]
where $i,j:\Nat$, $i\neq j$ and $i,j < n$. Note that we could remove
$\mathit{id}$ from the language and drop the $i\neq j$ condition so that
$\mathit{swap}~j~j$ would represent the identity permutation.

While we are mainly interested in typed data, it is always worthwhile
to first start investigating the untyped setting first, as it gives us
insights into what the \emph{purely operational} view of the theory
would be.  This tells us that permutations are an inescapable part of
the fabric of reversible computing.  However as permutations are
untyped, and act on the canonicalized version of $n$-element sets
(i.e. those sets where all the structure has been forgotten), these
are a rather pale shadow of the rich tapestry of
information-perserving transformation of structured data which we
investigate next.
